---
title: "Series01-RStudio"
---

### Analysis Plan

1.  Import the data and create the data frame required for the analysis.
2.  Display the data table and check that it was imported correctly.
3.  Check the structure of the data and the type of each variable.
4.  Obtain the main descriptive statistics and summary of the data.
5.  Interpret the summary and explain why it differs between columns.
6.  Check and clean the data before continuing with further analysis.
7.  Visualize the distribution of student weight using:

- a boxplot,
- a histogram,
- a density plot.

8.  Explain and label the x- and y-axes of each visualization, including the units.
9.  Use a boxplot to further examine and summarize the distribution of the data.
10. Compare the weight distributions of male and female students.
11. Choose and perform an appropriate statistical test to determine whether male and female weights are significantly different.
12. Interpret the statistical result and compare it with the visual evidence.

### Step 1 — Import the data and create the data frame

```{r Load libraries}

# Load tidyverse library, collection of packages for data manipulation and visualization.
library(tidyverse) 

# Create a dataframe by reading the CSV file from the "data" folder.

data_students <- read_csv("data/IEAP-2025 - Sheet1.csv") 

# <- means "store the result in data" We are storing the imported dataset in an object called "data".

# read_csv() belongs to the readr package, which is included in tidyverse.

```

### Step 2 — Display the data table and check the import

```{r Display the dataset}

# Display the dataset in the output (rows and columns)
data_students

# Open the complete dataset in RStudio's Data Viewer.
View(data_students) 
```

### Step 3 — Check the structure and variable types

```{r Check the dataset structure}

# use glimpse() before doing the analysis to check the structure of the data

glimpse(data) 

# <chr> = character → text, <dbl> = double → numerical value, <int> = integer → whole numerical value, <lgl> = logical → TRUE or FALSE

# YearsInPython appears as <chr> even though it should be numeric, which suggests an incorrectly formatted value. This is why we use glipse()
```

### Step 4 - Obtain descriptive statistics summary

```{r Obtain summary of students}

# Show the main summary statistics for every variable in the dataset.

summary(data_students)
```

### Step 5 - Interpret the summary

***Why is the summary different from one column to the next?***

- For **numerical variables** such as `BornYear`, `Weight`, `Height`, `YearsInR`, and `YearsInResearch`, R calculates descriptive statistics such as the minimum, quartiles, median, mean, and maximum.

- For **character/text variables** such as `Name`, `Firstname`, `Gender`, `City`, and `Lab`, these numerical statistics do not make sense, so R instead gives information such as the number of values, number of unique values, blank values, and character length.

- In this dataset, `YearsInPython` is also being treated as a character variable, which suggests that at least one value is not formatted as a number correctly. The data is written in the file with a comma (,) instead of (.).

### Step 6 - Check and clean the data

In this dataset, `YearsInPython` appears to have a formatting issue, but because the analysis below focuses on the `Weight` variable and `Weight` is already correctly stored as numerical data, no additional cleaning is required for the next steps.

### Step 7 - Visualize the biometric data

##### Figure 1 — Distribution of student weight boxplot

```{r Create a boxplot of student weight distribution}

# Create a horizontal boxplot of student weight using ggplot2.

ggplot(data_students, aes(x = Weight)) + # ggplot start/create a graph
  
  geom_boxplot() + # geometric object (geom) tells ggplot the type of visual representation.
  
  xlab("Weight (kg)") + # label the x-axis
  
  ggtitle("Figure 1: Distribution of Student Weight Boxplot") + # ggtitle adds the title above the graph
  
  theme_minimal() + # Use a clean visual theme
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"), # hjust centers the text
    axis.title.y = element_blank(), # leave y-axis title label blank
    axis.text.y = element_blank(), # leave y-axis label blank
    axis.ticks.y = element_blank() # leave y-axis ticks blank
  )
  
# data_students = data frame to use
# aes() = aesthetics mapping tells R which variables go where. x = Weight = put the Weight column on the x-axis
# + adds an additional layer to the plot
# theme() controls style of ggplot
# plot.title specifies which part of the graph you want to change
# element_text() tells ggplot that you want to change the formatting of text.
# hjust = horizontal justification (0=left, 0.5= center, 1= right)
  
```

> *Figure 1: Boxplot of Student Weight Distribution. The x-axis represents student weight in kilograms (kg). The boxplot summarizes the median, quartiles, and overall spread of student weight. The y-axis does not represent a measured variable and is therefore omitted.*

##### **Figure 2 — Distribution of student weight histogram**

```{r Create a histogram of student weight distributio}

ggplot(data_students, aes(x = Weight)) + # Start the graph and put Weight on the x-axis
  
  geom_histogram( breaks = seq (50, 85, by = 5), colour = "black", fill = "grey") + # Create a histogram by grouping weights into intervals (bins) at 5 kg intervals. Add black border and grey filling
  
  xlab("Weight (kg)") + # Label the x-axis with the variable and unit
  ylab("Frequency") + # Label the y-axis as the number of students in each weight interval
  
  ggtitle("Figure 2: Distribution of Student Weight Histogram") + # Add the figure title
  
  theme_minimal() + # Use a clean visual theme
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) # Center and bold the title

```

> *Figure 2: Histogram of Student Weight Distribution. The x-axis represents student weight in kilograms (kg), and the y-axis represents frequency, meaning the number of students falling within each 5 kg weight interval.*

##### Figure 3 — Distribution of student weight density plot

```{r Create a density plot of student weight distribution}

# Create a variable that stores the density estimate calculated from the Weight column.
weight_density <- density(data_students$Weight)

# Plot the density curve.
plot(weight_density,
     main = "Figure 3: Distribution of Student Weight Density Plot", # Title
     xlab = "Weight (kg)", # X-axis label and unit
     ylab = "Density") # Y-axis label
```

> *Figure 3: Density Plot of Student Weight Distribution. The x-axis represents student weight in kilograms (kg). The y-axis represents density, which is a smoothed estimate of how concentrated the student weight values are across the weight range. Higher parts of the curve indicate weights where more observations are concentrated, while lower parts indicate fewer observations. The total area under the density curve equals 1.*

### Step 8 — Use a boxplot to summarize the distribution

### Step 9 — Compare male and female weight distributions

### Step 10 — Perform a statistical test

### Step 11 — Interpret the results
