---
title: "Series01-RStudio"
---

### Analysis Plan

1.  Import the data and create the data frame required for the analysis.
2.  Display the data table and check that it was imported correctly.
3.  Check the structure of the data and the type of each variable.
4.  Obtain the main descriptive statistics and summary of the data.
5.  Interpret the summary and explain why it differs between columns.
6.  Check and clean the data before continuing with further analysis.
7.  Visualize the distribution of student weight using:

- a boxplot,
- a histogram,
- a density plot.

8.  Explain and label the x- and y-axes of each visualization, including the units.
9.  Use a boxplot to further examine and summarize the distribution of the data.
10. Compare the weight distributions of male and female students.
11. Choose and perform an appropriate statistical test to determine whether male and female weights are significantly different.
12. Interpret the statistical result and compare it with the visual evidence.

### Step 1 — Import the data and create the data frame

```{r Load libraries}

# Load tidyverse library, collection of packages for data manipulation and visualization.
library(tidyverse) 

# Create a dataframe by reading the CSV file from the "data" folder.

data_students <- read_csv("data/IEAP-2025 - Sheet1.csv") 

# <- means "store the result in data" We are storing the imported dataset in an object called "data".

# read_csv() belongs to the readr package, which is included in tidyverse.

```

### Step 2 — Display the data table and check the import

```{r Display the dataset}

# Display the dataset in the output (rows and columns)
data_students

# Open the complete dataset in RStudio's Data Viewer.
View(data_students) 
```

### Step 3 — Check the structure and variable types

```{r Check the dataset structure}

# use glimpse() before doing the analysis to check the structure of the data

glimpse(data) 

# <chr> = character → text, <dbl> = double → numerical value, <int> = integer → whole numerical value, <lgl> = logical → TRUE or FALSE

# YearsInPython appears as <chr> even though it should be numeric, which suggests an incorrectly formatted value. This is why we use glipse()
```

### Step 4 - Obtain descriptive statistics summary

```{r Obtain summary of students}

# Show the main summary statistics for every variable in the dataset.

summary(data_students)
```

### Step 5 - Interpret the summary

***Why is the summary different from one column to the next?***

- For **numerical variables** such as `BornYear`, `Weight`, `Height`, `YearsInR`, and `YearsInResearch`, R calculates descriptive statistics such as the minimum, quartiles, median, mean, and maximum.

- For **character/text variables** such as `Name`, `Firstname`, `Gender`, `City`, and `Lab`, these numerical statistics do not make sense, so R instead gives information such as the number of values, number of unique values, blank values, and character length.

- In this dataset, `YearsInPython` is also being treated as a character variable, which suggests that at least one value is not formatted as a number correctly. The data is written in the file with a comma (,) instead of (.).

### Step 6 - Check and clean the data

In this dataset, `YearsInPython` appears to have a formatting issue, but because the analysis below focuses on the `Weight` variable and `Weight` is already correctly stored as numerical data, no additional cleaning is required for the next steps.

### Step 7 - Visualize the biometric data

##### Figure 1 — Distribution of student weight boxplot

```{r Create a boxplot of student weight distribution}

# Create a horizontal boxplot of student weight using ggplot2.

ggplot(data_students, aes(x = Weight)) + # ggplot start/create a graph
  
  geom_boxplot() + # geometric object (geom) tells ggplot the type of visual representation.
  
  xlab("Weight (kg)") + # label the x-axis
  
  ggtitle("Figure 1: Distribution of Student Weight Boxplot") + # ggtitle adds the title above the graph
  
  theme_minimal() + # Use a clean visual theme
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"), # hjust centers the text
    axis.title.y = element_blank(), # leave y-axis title label blank
    axis.text.y = element_blank(), # leave y-axis label blank
    axis.ticks.y = element_blank() # leave y-axis ticks blank
  )
  
# data_students = data frame to use
# aes() = aesthetics mapping tells R which variables go where. x = Weight = put the Weight column on the x-axis
# + adds an additional layer to the plot
# theme() controls style of ggplot
# plot.title specifies which part of the graph you want to change
# element_text() tells ggplot that you want to change the formatting of text.
# hjust = horizontal justification (0=left, 0.5= center, 1= right)
  
```

> *Figure 1: Boxplot of Student Weight Distribution. The x-axis represents student weight in kilograms (kg). The boxplot summarizes the median, quartiles, and overall spread of student weight. The y-axis does not represent a measured variable and is therefore omitted.*

##### **Figure 2 — Distribution of student weight histogram**

```{r Create a histogram of student weight distributio}

ggplot(data_students, aes(x = Weight)) + # Start the graph and put Weight on the x-axis
  
  geom_histogram( breaks = seq (50, 85, by = 5), colour = "black", fill = "grey") + # Create a histogram by grouping weights into intervals (bins) at 5 kg intervals. Add black border and grey filling
  
  xlab("Weight (kg)") + # Label the x-axis with the variable and unit
  ylab("Frequency") + # Label the y-axis as the number of students in each weight interval
  
  ggtitle("Figure 2: Distribution of Student Weight Histogram") + # Add the figure title
  
  theme_minimal() + # Use a clean visual theme
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) # Center and bold the title

```

> *Figure 2: Histogram of Student Weight Distribution. The x-axis represents student weight in kilograms (kg), and the y-axis represents frequency, meaning the number of students falling within each 5 kg weight interval.*

##### Figure 3 — Distribution of student weight density plot

```{r Create a density plot of student weight distribution}

# Create a variable that stores the density estimate calculated from the Weight column.
weight_density <- density(data_students$Weight)

# Plot the density curve.
plot(weight_density,
     main = "Figure 3: Distribution of Student Weight Density Plot", # Title
     xlab = "Weight (kg)", # X-axis label and unit
     ylab = "Density") # Y-axis label
```

> *Figure 3: Density Plot of Student Weight Distribution. The x-axis represents student weight in kilograms (kg). The y-axis represents density, which is a smoothed estimate of how concentrated the student weight values are across the weight range. Higher parts of the curve indicate weights where more observations are concentrated, while lower parts indicate fewer observations. The total area under the density curve equals 1.*

### Step 8 — Why use a boxplot to summarize the distribution?

**a) Main elements of a boxplot?**

A boxplot is a graphical representation of a distribution based mainly on the median and quartiles. Its main elements are

- **Q1:** the value below which 25% of the observations fall.
- **Q2:** the middle value of the dataset, representing the 50th percentile.
- **Q3:** the value below which 75% of the observations fall.
- **Box:** extends from Q1 to Q3 and therefore contains the middle 50% of the observations.
- **Interquartile range (IQR):** The **IQR (interquartile range)** is calculated as `Q3 − Q1` and represents the spread of the middle 50% of the data. It is useful because it shows how variable the central part of the distribution is without being strongly affected by extreme values.
- **1.5 × IQR rule** : used to identify possible outliers: values below `Q1 − 1.5 × IQR` or above `Q3 + 1.5 × IQR` are considered potential outliers, while the whiskers extend to the most extreme values that are still within these limits.
- **Whiskers:** The whiskers extend from the box to the smallest and largest values that are still considered part of the normal data, while values beyond the whiskers are shown separately as possible outliers.
- **Outliers:** observations beyond the whiskers.

**b) Why is a boxplot the starting point of choice in summarizing an unknown distribution?**

- A boxplot is useful at the beginning because it quickly shows several important things about data before you know what the distribution looks like. It lets you see:

  - the center of the data through the median

  - the spread through the quartiles and IQR,

  - whether the data looks symmetric or skewed

  - whether there are any possible outliers.

  - It is not affected by extreme values because it is based on the median and IQR rather than the mean and standard deviation.

### Step 9 — Reproduce the boxplot to summarize the distribution

```{r First calculate the mean and its 95% confidence interval:}

# Calculate the number of students, mean weight and standard error.

# length counts how many weight values there are in the column.
n_students <- length(data_students$Weight) 

mean_weight <- mean(data_students$Weight) # Calculate the mean student weight.

# standard error = standard deviation / √number of students
se_weight <- sd(data_students$Weight) / sqrt(n_students) 

# Calculate the lower and upper limits of the 95% confidence interval of the mean.

ci_margin <- qt(0.975, df = n_students - 1) * se_weight # Calculate how far the 95% confidence interval extends above and below the mean.

ci_lower <- mean_weight - ci_margin # Calculate the lower limit of the 95% confidence interval.

ci_upper <- mean_weight + ci_margin # Calculate the upper limit of the 95% confidence interval.
```

```{r Now plot the combined graphs}

ggplot() + # Start an empty ggplot because different parts of the graph will be positioned separately.
  geom_jitter(data = data_students, aes(x = 1, y = Weight), width = 0.25) + # geom_jitter() plots each students weight as dots. x = 1 puts all the dots roughly on the left side.
  geom_boxplot(data = data_students, aes(x = 2, y = Weight), width = 0.6) + # Summarize the weight distribution with a boxplot. x = 2 places it in the middle.
  geom_errorbar(aes(x = 3, ymin = ci_lower, ymax = ci_upper), width = 0.15) + # Show the 95% confidence interval of the mean. x = 3 places it on the right.
  geom_point(aes(x = 3, y = mean_weight), size = 3) + # Show the mean weight
  scale_x_continuous(breaks = NULL) + # Remove x-axis numbers because they only position the three elements
  xlab("") + # Remove the x-axis label because there is no x-axis.
  ylab("Weight (kg)") + # Label the y-axis as student weight in kilograms.
  ggtitle("Figure 4: Summary of Student Weight Distribution") +  # Add the title.
  theme_minimal() + # Apply a clean visual style.
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) # Center the title and make it bold.
```

> > *Figure 4: The three representations are shown together because they provide complementary information about the same data: the individual points show the raw observations, the boxplot summarizes the distribution, and the mean with its 95% confidence interval shows the estimated average and the uncertainty around that estimate.*

**a) What is a confidence interval?**

- A **confidence interval (CI)** gives a range around an estimate, in this case, around the sample mean weight—to show how precise that estimate is.

- A narrow CI means the mean is estimated relatively precisely; a wide CI means there is more uncertainty. The width depends especially on sample size and how variable the data are.

### Step 10 — Compare male and female weight distributions

**a) Is the distribution for males and females similar?**

Generate a histogram:

```{r Create a histogram to show the weight distribution by gender}

ggplot(data_students, aes(x = Weight, fill = Gender)) + # Put Weight on the x-axis and and use different fills for Gender
  geom_histogram(breaks = seq(50, 85, by = 5), position = "identity", alpha = 0.6, colour = "black") + # Create 5 kg bins intervals
# position = identity means draw the female and male on top of each other when they occupy the same weight interval.
# alpha controls transparency ( 1= completley solid)
  scale_fill_manual(values = c("F" = "red", "M" = "blue")) + # Red for females and blue for males
  
  
  xlab("Weight (kg)") + # Label the x-axis and include the unit
  ylab("Frequency (students)") + # Show the number of students in each weight interval
  ggtitle("Figure 5: Weight Distribution by Gender") + # Add the figure title
  theme_minimal() + # Use a clean visual theme
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) # Center and bold the title
```

**a) Is the histogram the best way to visualize this data for this question?**

- With a small sample, a histogram can be confusing because the shape of the graph can change a lot depending on how the bins are chosen.
- The same data can look more separated, more grouped, or more uneven just because the weight ranges are divided differently.
- Since there are only a few students in each gender group, one or two values can noticeably change the height of a bar, so the graph may make the differences between males and females look bigger than they really are.

### Step 11 — **Is the weight of males/females significantly different?**

***Is the weight of males/females significantly different?***

**a) What visualization is best to answer this question?**

- A boxplot with the individual data points is better than the histogram as it lets you directly compare the median and spread of male and female weights, while the individual points show the actual observations and make the small sample size visible.

**b) What statistical test is best to answer this question?**

- Wilcoxon rank-sum test, also called the Mann–Whitney U test

  - We have two independent groups (male and female) and a relatively small sample.

  - It does not require the weight data to follow a normal distribution.

-  A p-value below 0.05 indicates that the difference between the groups is statistically significant.

### Step 12 — Important Concepts

#### Specific Terms

> **Boxplot:** summarizes numerical data using the median, quartiles, whiskers, and possible outliers.

> **IQR (Interquartile Range):** `Q3 - Q1`; shows the spread of the middle 50% of the data.

> **Whiskers:** extend to the smallest and largest values that are not considered outliers.

> **Outliers:** values below `Q1 - 1.5 × IQR` or above `Q3 + 1.5 × IQR`.

> **Distribution:** shows how values are spread and concentrated in a dataset.

> **Histogram:** groups numerical values into bins and shows the frequency in each bin.

> **Bin width:** size of each interval in a histogram; changing it can change how the distribution looks.

> **Density plot:** smooth estimate of where values are concentrated. The total area under the curve equals 1.

> **Mean:** average of all values.

> **Standard deviation (SD):** shows how spread out individual values are around the mean.

> **Standard error (SE):** shows how precise the sample mean is as an estimate of the population mean.

- `SE = SD / √n`

> **Confidence interval (CI):** range showing the uncertainty around an estimate. A narrower CI means a more precise estimate.

> **Wilcoxon rank-sum test:** compares a numerical variable between two independent groups without assuming normality.

> **P-value:** helps determine whether the observed difference is statistically significant.

- `p < 0.05` is usually considered statistically significant.

> **Effect size:** shows how large the difference between groups is. `- r ≈ 0.1` = small, `- r ≈ 0.3` = medium, - r ≥ 0.5\` = large.

> **Independent groups:** separate groups where each observation belongs to only one group.

```         
```
