---
title: "Series01-RStudio"
---

### Analysis Plan

1.  Import data and create data frame
2.  Check that the tabel was correctly uploaded
3.  Check the structure of the data
4.  Get the descriptive statistics of the data
5.  Interpret the summary of the data
6.  

### Step 1 — Import the data and create data frame required for the data analysis

```{r Load libraries}
# Load tidyverse library, collection of packages for data manipulation and visualization.
library(tidyverse) # read_csv() belongs to the readr package, which is included in tidyverse.

# Create datafram by reading the CSV file from the "data" folder.
data_students <- read_csv("data/IEAP-2025 - Sheet1.csv") # <- means "store the result in data" We are storing the imported dataset in an object called "data".

```

### Step 2 — Check that the table was imported correctly

```{r}
# Display the dataset in the output (rows and columns)
data_students

View(data_students) #A tibble is the tidyverse's modern version of a data frame.
```

### Step 3 — Check the structure of the data

```{r}
# use glimpse() before doing the analysis to check the structure of the data

glimpse(data) # <chr> = character → text, <dbl> = double → numerical value, <int> = integer → whole numerical value, <lgl> = logical → TRUE or FALSE

# YearsInPython <chr> should be numeric and this is why we use glimpse()
```

### Step 4 - Summary of Student Dataframe

```{r}
summary(data_students)
```

Why is the summary different from one column to the next?

For **numerical variables** such as `BornYear`, `Weight`, `Height`, `YearsInR`, and `YearsInResearch`, R calculates descriptive statistics such as the **minimum, quartiles, median, mean, and maximum**.

For **character/text variables** such as `Name`, `Firstname`, `Gender`, `City`, and `Lab`, these numerical statistics do not make sense, so R instead gives information such as the **number of values, number of unique values, blank values, and character length**.

In this dataset, `YearsInPython` is also being treated as a character variable, which suggests that at least one value is not formatted as a number correctly. The data is written in the file with a comma (,) instead of (.).

### Step 5 - 
